---
title: "Cross-Dataset Prediction of a Temozolomide-Response-Associated Biomarker in Glioblastoma"
author: "Saidat Bello"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

# Question 1

## Study aim

This analysis addresses the following project question:

> Can a predictor of patient response be developed that works across different datasets?

This analysis addresses the project question indirectly by predicting a
temozolomide-response-associated biomarker rather than observed patient
response. A gene-expression classifier is developed in the TCGA glioblastoma
cohort and externally evaluated in the CPTAC glioblastoma cohort. The outcome
is MGMT promoter methylation, which is associated with temozolomide benefit
but is not itself a direct measurement of treatment response.

The specific analysis question is:

> Can a gene-expression predictor of an MGMT methylation biomarker associated
> with temozolomide response transfer from TCGA to CPTAC?

The analysis prioritises features measured in both cohorts. This provides a practical compromise between model completeness and cross-dataset portability.

# 1. Analysis setup

```{r}
#| label: setup
#| include: false

required_packages <- c(
  "data.table",
  "dplyr",
  "tidyr",
  "stringr",
  "tibble",
  "janitor",
  "glmnet",
  "pROC",
  "ggplot2",
  "knitr"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Install the following packages before rendering: ",
      paste(missing_packages, collapse = ", ")
    )
  )
}

library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(janitor)
library(glmnet)
library(pROC)
library(ggplot2)
library(knitr)

set.seed(123)
```

```{r}
#| label: file-paths
#| include: false

project_directory <- "/Users/saidatbello/Documents/AI in P - final project"

tcga_patient_file <- file.path(
  project_directory,
  "gbm_tcga_pan_can_atlas_2018:data_clinical_patient.txt"
)
tcga_sample_file <- file.path(project_directory, "data_clinical_sample.txt")
tcga_expression_file <- file.path(
  project_directory,
  "data_mrna_seq_v2_rsem.txt"
)
tcga_methylation_file <- file.path(
  project_directory,
  "data_methylation_hm450.txt"
)

cptac_patient_file <- file.path(
  project_directory,
  "gbm_cptac_2021:data_clinical_patient.txt"
)
cptac_sample_file <- file.path(project_directory, "data_clinical_sample1.txt")
cptac_expression_file <- file.path(
  project_directory,
  "data_mrna_seq_fpkm.txt"
)
cptac_methylation_file <- file.path(
  project_directory,
  "data_methylation_epic.txt"
)

required_files <- c(
  tcga_patient_file,
  tcga_sample_file,
  tcga_expression_file,
  tcga_methylation_file,
  cptac_patient_file,
  cptac_sample_file,
  cptac_expression_file,
  cptac_methylation_file
)

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    paste0(
      "The following files were not found: ",
      paste(missing_files, collapse = ", ")
    )
  )
}
```

# 2. Data sources and cohort preparation

TCGA was used as the model-development cohort and CPTAC was used as the independent external-validation cohort. Patient and sample files were used to confirm identifiers and retain one tumour sample per patient.

```{r}
#| label: load-clinical

read_cbio_clinical <- function(path) {
  readr::read_tsv(
    path,
    comment = "#",
    show_col_types = FALSE,
    na = c("", "NA", "NaN")
  ) |>
    clean_names()
}

tcga_patient <- read_cbio_clinical(tcga_patient_file)
tcga_sample <- read_cbio_clinical(tcga_sample_file)

cptac_patient <- read_cbio_clinical(cptac_patient_file)
cptac_sample <- read_cbio_clinical(cptac_sample_file)
```

```{r}
#| label: select-one-sample

tcga_sample_primary <- tcga_sample |>
  mutate(
    sample_priority = case_when(
      str_to_lower(sample_type) == "primary" ~ 1,
      str_detect(str_to_lower(tumor_type), "untreated") ~ 1,
      TRUE ~ 2
    )
  ) |>
  arrange(patient_id, sample_priority, sample_id) |>
  distinct(patient_id, .keep_all = TRUE) |>
  select(-sample_priority)

cptac_sample_primary <- cptac_sample |>
  arrange(patient_id, sample_id) |>
  distinct(patient_id, .keep_all = TRUE)
```

```{r}
#| label: cohort-summary

cohort_summary <- tibble(
  cohort = c("TCGA", "CPTAC"),
  patient_records = c(nrow(tcga_patient), nrow(cptac_patient)),
  sample_records = c(nrow(tcga_sample), nrow(cptac_sample)),
  unique_patients_after_selection = c(
    n_distinct(tcga_sample_primary$patient_id),
    n_distinct(cptac_sample_primary$patient_id)
  )
)

kable(
  cohort_summary,
  caption = "Clinical cohort summary before molecular-data matching"
)
```

# 3. MGMT promoter methylation outcome

The supplied TCGA file is probe-level, whereas the supplied CPTAC file is
gene-indexed and does not retain CpG probe identifiers. A comparable
gene-level MGMT methylation score was therefore constructed in both cohorts
by averaging all rows annotated to `MGMT` for each selected tumour sample.
This is a study-specific molecular outcome and is not the clinical
MGMT-STP27 classifier.

Only MGMT-annotated rows are extracted from each large methylation file,
avoiding the need to load the complete matrices into memory.

```{r}
#| label: methylation-functions
#| include: false

extract_mgmt_rows <- function(file_path, cohort) {
  header <- readLines(file_path, n = 1, warn = FALSE)

  grep_pattern <- if (cohort == "TCGA") {
    "\tMGMT\t"
  } else {
    "^MGMT\t"
  }
  temp_file <- tempfile(fileext = ".txt")

  system2(
    command = "grep",
    args = c("-E", shQuote(grep_pattern), shQuote(file_path)),
    stdout = temp_file,
    stderr = FALSE
  )

  matching_rows <- readLines(temp_file, warn = FALSE)
  unlink(temp_file)

  if (length(matching_rows) == 0) {
    stop(
      paste0(
        "No MGMT-annotated methylation rows were found in ",
        basename(file_path),
        ". Check the file structure."
      )
    )
  }

  fread(
    text = paste(c(header, matching_rows), collapse = "\n"),
    sep = "\t",
    na.strings = c("", "NA", "NaN"),
    data.table = FALSE,
    check.names = FALSE
  )
}

methylation_gene_score <- function(data, cohort, allowed_samples) {
  data |>
    select(any_of(allowed_samples)) |>
    pivot_longer(
      cols = everything(),
      names_to = "sample_id",
      values_to = "beta_value"
    ) |>
    mutate(
      beta_value = suppressWarnings(as.numeric(beta_value)),
      cohort = cohort
    ) |>
    filter(!is.na(beta_value), sample_id %in% allowed_samples) |>
    group_by(cohort, sample_id) |>
    summarise(mgmt_beta = mean(beta_value), .groups = "drop")
}

derive_patient_id <- function(sample_id, cohort) {
  if (cohort == "TCGA") {
    substr(sample_id, 1, 12)
  } else {
    sample_id
  }
}
```

```{r}
#| label: extract-mgmt-methylation

tcga_mgmt_raw <- extract_mgmt_rows(
  tcga_methylation_file,
  cohort = "TCGA"
)

cptac_mgmt_raw <- extract_mgmt_rows(
  cptac_methylation_file,
  cohort = "CPTAC"
)

tcga_mgmt <- methylation_gene_score(
  tcga_mgmt_raw,
  cohort = "TCGA",
  allowed_samples = tcga_sample_primary$sample_id
) |>
  mutate(
    patient_id = derive_patient_id(sample_id, "TCGA")
  )

cptac_mgmt <- methylation_gene_score(
  cptac_mgmt_raw,
  cohort = "CPTAC",
  allowed_samples = cptac_sample_primary$sample_id
) |>
  mutate(
    patient_id = derive_patient_id(sample_id, "CPTAC")
  )
```

## 3.1 Defining methylated and unmethylated groups

A two-group clustering model was fitted to the continuous, gene-level TCGA
MGMT methylation scores. The midpoint between the two cluster centres was
retained as the fixed classification threshold and applied unchanged to
CPTAC.

```{r}
#| label: define-mgmt-status

tcga_kmeans <- kmeans(
  tcga_mgmt$mgmt_beta,
  centers = 2,
  nstart = 50
)

cluster_centres <- sort(as.numeric(tcga_kmeans$centers))
mgmt_cutpoint <- mean(cluster_centres)

tcga_mgmt <- tcga_mgmt |>
  mutate(
    mgmt_status = if_else(
      mgmt_beta >= mgmt_cutpoint,
      "Methylated",
      "Unmethylated"
    )
  )

cptac_mgmt <- cptac_mgmt |>
  mutate(
    mgmt_status = if_else(
      mgmt_beta >= mgmt_cutpoint,
      "Methylated",
      "Unmethylated"
    )
  )
```

```{r}
#| label: mgmt-cutpoint-output

tibble(
  lower_cluster_centre = cluster_centres[1],
  upper_cluster_centre = cluster_centres[2],
  fixed_cutpoint = mgmt_cutpoint
) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable(
    caption = "TCGA-derived MGMT methylation classification threshold"
  )
```

```{r}
#| label: mgmt-class-distribution

bind_rows(
  tcga_mgmt |> select(cohort, mgmt_status),
  cptac_mgmt |> select(cohort, mgmt_status)
) |>
  count(cohort, mgmt_status) |>
  group_by(cohort) |>
  mutate(percent = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  kable(
    caption = "MGMT promoter methylation class distribution"
  )
```

```{r}
#| label: mgmt-distribution-plot
#| fig-cap: "Distribution of the gene-level MGMT methylation score in TCGA and CPTAC."

bind_rows(
  tcga_mgmt |> select(cohort, mgmt_beta, mgmt_status),
  cptac_mgmt |> select(cohort, mgmt_beta, mgmt_status)
) |>
  ggplot(aes(x = mgmt_beta, fill = mgmt_status)) +
  geom_histogram(bins = 25, alpha = 0.65, position = "identity") +
  geom_vline(
    xintercept = mgmt_cutpoint,
    linetype = "dashed"
  ) +
  facet_wrap(~ cohort, scales = "free_y") +
  labs(
    x = "Mean MGMT-annotated beta value",
    y = "Number of samples",
    fill = "MGMT status"
  ) +
  theme_minimal()
```

# 4. Gene-expression preprocessing

The TCGA and CPTAC expression matrices use different quantification methods. Both matrices were therefore log-transformed. Only gene symbols present in both cohorts were retained.

```{r}
#| label: expression-functions
#| include: false

read_expression_matrix <- function(path, cohort, allowed_samples) {
  expression_data <- fread(
    path,
    sep = "\t",
    na.strings = c("", "NA", "NaN"),
    data.table = FALSE,
    check.names = FALSE
  )

  gene_column <- names(expression_data)[
    str_detect(
      str_to_lower(names(expression_data)),
      "hugo_symbol|gene_symbol|symbol"
    )
  ][1]

  if (is.na(gene_column)) {
    gene_column <- names(expression_data)[1]
  }

  expression_data <- expression_data |>
    rename(gene = all_of(gene_column)) |>
    mutate(gene = as.character(gene)) |>
    filter(!is.na(gene), gene != "")

  expression_data <- expression_data |>
    select(gene, any_of(allowed_samples))

  expression_long <- expression_data |>
    pivot_longer(
      cols = -gene,
      names_to = "sample_id",
      values_to = "expression"
    ) |>
    mutate(
      expression = suppressWarnings(as.numeric(expression)),
      patient_id = derive_patient_id(sample_id, cohort)
    ) |>
    filter(!is.na(expression)) |>
    group_by(patient_id, gene) |>
    summarise(expression = mean(expression), .groups = "drop")

  expression_long
}
```

```{r}
#| label: load-expression

tcga_expression_long <- read_expression_matrix(
  tcga_expression_file,
  cohort = "TCGA",
  allowed_samples = tcga_sample_primary$sample_id
)

cptac_expression_long <- read_expression_matrix(
  cptac_expression_file,
  cohort = "CPTAC",
  allowed_samples = cptac_sample_primary$sample_id
)
```

```{r}
#| label: common-genes

common_genes <- intersect(
  unique(tcga_expression_long$gene),
  unique(cptac_expression_long$gene)
)

expression_overlap <- tibble(
  measure = c(
    "TCGA genes",
    "CPTAC genes",
    "Genes present in both cohorts"
  ),
  number = c(
    n_distinct(tcga_expression_long$gene),
    n_distinct(cptac_expression_long$gene),
    length(common_genes)
  )
)

kable(
  expression_overlap,
  caption = "Gene-expression feature availability across cohorts"
)
```

## 4.1 Feature filtering

Genes were filtered using TCGA only. Genes with excessive missingness or near-zero variance were removed. The 500 most variable genes among those also measured in CPTAC were retained as candidate predictors.

```{r}
#| label: select-features

tcga_candidate <- tcga_expression_long |>
  filter(gene %in% common_genes) |>
  group_by(gene) |>
  summarise(
    variance = var(log2(expression + 1), na.rm = TRUE),
    non_missing = sum(!is.na(expression)),
    .groups = "drop"
  ) |>
  filter(
    is.finite(variance),
    variance > 0,
    non_missing >= 0.8 * n_distinct(tcga_expression_long$patient_id)
  ) |>
  arrange(desc(variance)) |>
  slice_head(n = 500)

selected_genes <- tcga_candidate$gene
```

```{r}
#| label: selected-feature-summary

tibble(
  candidate_genes_available_in_both_cohorts = length(common_genes),
  genes_entering_model_development = length(selected_genes)
) |>
  kable(
    caption = "Feature-filtering summary"
  )
```

# 5. Construction of modelling datasets

```{r}
#| label: create-expression-wide
#| include: false

make_expression_wide <- function(data, genes) {
  data |>
    filter(gene %in% genes) |>
    mutate(expression = log2(expression + 1)) |>
    select(patient_id, gene, expression) |>
    pivot_wider(
      names_from = gene,
      values_from = expression,
      values_fn = mean
    )
}

tcga_expression_wide <- make_expression_wide(
  tcga_expression_long,
  selected_genes
)

cptac_expression_wide <- make_expression_wide(
  cptac_expression_long,
  selected_genes
)

tcga_model_data <- tcga_mgmt |>
  select(patient_id, mgmt_status, mgmt_beta) |>
  inner_join(tcga_expression_wide, by = "patient_id") |>
  distinct(patient_id, .keep_all = TRUE)

cptac_model_data <- cptac_mgmt |>
  select(patient_id, mgmt_status, mgmt_beta) |>
  inner_join(cptac_expression_wide, by = "patient_id") |>
  distinct(patient_id, .keep_all = TRUE)
```

```{r}
#| label: matched-cohort-summary

tibble(
  cohort = c("TCGA development", "CPTAC validation"),
  matched_patients = c(
    nrow(tcga_model_data),
    nrow(cptac_model_data)
  ),
  methylated = c(
    sum(tcga_model_data$mgmt_status == "Methylated"),
    sum(cptac_model_data$mgmt_status == "Methylated")
  ),
  unmethylated = c(
    sum(tcga_model_data$mgmt_status == "Unmethylated"),
    sum(cptac_model_data$mgmt_status == "Unmethylated")
  )
) |>
  kable(
    caption = "Final matched cohorts used for model development and validation"
  )
```

# 6. Model development in TCGA

A LASSO logistic-regression model was used because it performs feature selection while limiting overfitting in high-dimensional molecular data.

The outcome was coded as:

- `1`: methylated
- `0`: unmethylated

Missing expression values were imputed using TCGA medians. TCGA means and standard deviations were then used to standardise both datasets, preventing information from CPTAC from influencing model development.

```{r}
#| label: prepare-model-matrices
#| include: false

feature_columns <- intersect(
  selected_genes,
  intersect(
    names(tcga_model_data),
    names(cptac_model_data)
  )
)

x_tcga_raw <- as.matrix(
  tcga_model_data[, feature_columns, drop = FALSE]
)

x_cptac_raw <- as.matrix(
  cptac_model_data[, feature_columns, drop = FALSE]
)

storage.mode(x_tcga_raw) <- "numeric"
storage.mode(x_cptac_raw) <- "numeric"

x_tcga_unprocessed <- x_tcga_raw

training_medians <- apply(
  x_tcga_raw,
  2,
  median,
  na.rm = TRUE
)

for (column in seq_len(ncol(x_tcga_raw))) {
  x_tcga_raw[is.na(x_tcga_raw[, column]), column] <-
    training_medians[column]

  x_cptac_raw[is.na(x_cptac_raw[, column]), column] <-
    training_medians[column]
}

training_means <- colMeans(x_tcga_raw)
training_sds <- apply(x_tcga_raw, 2, sd)

usable_columns <- is.finite(training_sds) & training_sds > 0

x_tcga_raw <- x_tcga_raw[, usable_columns, drop = FALSE]
x_cptac_raw <- x_cptac_raw[, usable_columns, drop = FALSE]

training_means <- training_means[usable_columns]
training_sds <- training_sds[usable_columns]

x_tcga <- sweep(x_tcga_raw, 2, training_means, "-")
x_tcga <- sweep(x_tcga, 2, training_sds, "/")

x_cptac <- sweep(x_cptac_raw, 2, training_means, "-")
x_cptac <- sweep(x_cptac, 2, training_sds, "/")

y_tcga <- ifelse(
  tcga_model_data$mgmt_status == "Methylated",
  1,
  0
)

y_cptac <- ifelse(
  cptac_model_data$mgmt_status == "Methylated",
  1,
  0
)
```

## 6.1 Internal cross-validation

Five-fold cross-validation was performed entirely within TCGA. Each held-out fold was predicted by a model that had not seen those patients.
Within every outer fold, missing-value imputation and standardisation were
estimated from that fold's training patients and then applied to its held-out
patients.

```{r}
#| label: internal-cross-validation

number_of_folds <- 5

fold_id <- integer(length(y_tcga))

for (class_value in unique(y_tcga)) {
  class_indices <- which(y_tcga == class_value)

  fold_id[class_indices] <- sample(
    rep(
      seq_len(number_of_folds),
      length.out = length(class_indices)
    )
  )
}

tcga_cv_probability <- rep(NA_real_, length(y_tcga))

for (fold in seq_len(number_of_folds)) {
  training_index <- fold_id != fold
  validation_index <- fold_id == fold

  fold_train <- x_tcga_unprocessed[training_index, , drop = FALSE]
  fold_validation <- x_tcga_unprocessed[validation_index, , drop = FALSE]

  fold_medians <- apply(fold_train, 2, median, na.rm = TRUE)
  fold_medians[!is.finite(fold_medians)] <- 0

  for (column in seq_len(ncol(fold_train))) {
    fold_train[is.na(fold_train[, column]), column] <-
      fold_medians[column]
    fold_validation[is.na(fold_validation[, column]), column] <-
      fold_medians[column]
  }

  fold_means <- colMeans(fold_train)
  fold_sds <- apply(fold_train, 2, sd)
  fold_sds[!is.finite(fold_sds) | fold_sds == 0] <- 1

  fold_train <- sweep(fold_train, 2, fold_means, "-")
  fold_train <- sweep(fold_train, 2, fold_sds, "/")
  fold_validation <- sweep(fold_validation, 2, fold_means, "-")
  fold_validation <- sweep(fold_validation, 2, fold_sds, "/")

  fold_model <- cv.glmnet(
    x = fold_train,
    y = y_tcga[training_index],
    family = "binomial",
    alpha = 1,
    nfolds = 5,
    type.measure = "deviance",
    standardize = FALSE
  )

  tcga_cv_probability[validation_index] <- as.numeric(
    predict(
      fold_model,
      newx = fold_validation,
      s = "lambda.min",
      type = "response"
    )
  )
}

tcga_cv_roc <- roc(
  response = y_tcga,
  predictor = tcga_cv_probability,
  quiet = TRUE
)
```

## 6.2 Final frozen model

The final model was fitted using all TCGA patients. The selected coefficients and TCGA preprocessing parameters were then frozen before CPTAC evaluation.

```{r}
#| label: fit-final-model

final_model <- cv.glmnet(
  x = x_tcga,
  y = y_tcga,
  family = "binomial",
  alpha = 1,
  nfolds = 10,
  type.measure = "deviance",
  standardize = FALSE
)

final_coefficients <- coef(
  final_model,
  s = "lambda.min"
)

selected_coefficient_table <- tibble(
  feature = rownames(final_coefficients),
  coefficient = as.numeric(final_coefficients)
) |>
  filter(
    feature != "(Intercept)",
    coefficient != 0
  ) |>
  arrange(desc(abs(coefficient)))
```

```{r}
#| label: coefficient-table

selected_coefficient_table |>
  mutate(coefficient = round(coefficient, 4)) |>
  kable(
    caption = "Non-zero genes selected by the final TCGA LASSO model"
  )
```

```{r}
#| label: coefficient-plot
#| fig-cap: "Genes retained in the final TCGA LASSO classifier."

selected_coefficient_table |>
  slice_head(n = 20) |>
  mutate(
    feature = reorder(feature, coefficient)
  ) |>
  ggplot(aes(x = feature, y = coefficient)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "LASSO coefficient"
  ) +
  theme_minimal()
```

# 7. Independent validation in CPTAC

The frozen TCGA model was applied to CPTAC without refitting coefficients, recalculating feature selection or using CPTAC outcome information.

```{r}
#| label: external-predictions

cptac_probability <- as.numeric(
  predict(
    final_model,
    newx = x_cptac,
    s = "lambda.min",
    type = "response"
  )
)

cptac_roc <- roc(
  response = y_cptac,
  predictor = cptac_probability,
  quiet = TRUE
)

# A prespecified probability threshold avoids optimistically selecting and
# evaluating a cutpoint using the same out-of-fold predictions.
classification_threshold <- 0.5

tcga_cv_class <- ifelse(
  tcga_cv_probability >= classification_threshold,
  1,
  0
)

cptac_class <- ifelse(
  cptac_probability >= classification_threshold,
  1,
  0
)
```

```{r}
#| label: metric-functions
#| include: false

classification_metrics <- function(truth, probability, predicted_class) {
  true_positive <- sum(truth == 1 & predicted_class == 1)
  true_negative <- sum(truth == 0 & predicted_class == 0)
  false_positive <- sum(truth == 0 & predicted_class == 1)
  false_negative <- sum(truth == 1 & predicted_class == 0)

  tibble(
    auc = as.numeric(auc(roc(truth, probability, quiet = TRUE))),
    accuracy = (true_positive + true_negative) / length(truth),
    sensitivity = ifelse(
      true_positive + false_negative == 0,
      NA_real_,
      true_positive / (true_positive + false_negative)
    ),
    specificity = ifelse(
      true_negative + false_positive == 0,
      NA_real_,
      true_negative / (true_negative + false_positive)
    ),
    positive_predictive_value = ifelse(
      true_positive + false_positive == 0,
      NA_real_,
      true_positive / (true_positive + false_positive)
    ),
    negative_predictive_value = ifelse(
      true_negative + false_negative == 0,
      NA_real_,
      true_negative / (true_negative + false_negative)
    )
  )
}
```

```{r}
#| label: performance-table

performance_table <- bind_rows(
  classification_metrics(
    y_tcga,
    tcga_cv_probability,
    tcga_cv_class
  ) |>
    mutate(cohort = "TCGA: five-fold cross-validation"),
  classification_metrics(
    y_cptac,
    cptac_probability,
    cptac_class
  ) |>
    mutate(cohort = "CPTAC: external validation")
) |>
  select(cohort, everything()) |>
  mutate(
    across(
      where(is.numeric),
      \(x) round(x, 3)
    )
  )

kable(
  performance_table,
  caption = "Internal and external model performance"
)
```

```{r}
#| label: roc-plot
#| fig-cap: "Receiver operating characteristic curves for TCGA internal cross-validation and CPTAC external validation."

roc_plot_data <- bind_rows(
  tibble(
    specificity = tcga_cv_roc$specificities,
    sensitivity = tcga_cv_roc$sensitivities,
    cohort = "TCGA cross-validation"
  ),
  tibble(
    specificity = cptac_roc$specificities,
    sensitivity = cptac_roc$sensitivities,
    cohort = "CPTAC external validation"
  )
)

ggplot(
  roc_plot_data,
  aes(
    x = 1 - specificity,
    y = sensitivity,
    linetype = cohort
  )
) +
  geom_line(linewidth = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dotted"
  ) +
  coord_equal() +
  labs(
    x = "1 - specificity",
    y = "Sensitivity",
    linetype = NULL
  ) +
  theme_minimal()
```

```{r}
#| label: confusion-matrices

tcga_confusion <- table(
  Predicted = ifelse(
    tcga_cv_class == 1,
    "Methylated",
    "Unmethylated"
  ),
  Observed = tcga_model_data$mgmt_status
)

cptac_confusion <- table(
  Predicted = ifelse(
    cptac_class == 1,
    "Methylated",
    "Unmethylated"
  ),
  Observed = cptac_model_data$mgmt_status
)

cat("TCGA cross-validation confusion matrix\n\n")
print(tcga_confusion)

cat("\n\nCPTAC external-validation confusion matrix\n\n")
print(cptac_confusion)
```

# 8. Results summary

```{r}
#| label: dynamic-results
#| echo: false
#| results: asis

tcga_auc_value <- round(as.numeric(auc(tcga_cv_roc)), 3)
cptac_auc_value <- round(as.numeric(auc(cptac_roc)), 3)
number_selected <- nrow(selected_coefficient_table)
cptac_predicted_methylated <- sum(cptac_class == 1)
cptac_specificity_value <- round(
  sum(y_cptac == 0 & cptac_class == 0) / sum(y_cptac == 0),
  3
)

cat(
  paste0(
    "After matching methylation and expression data, the development cohort included ",
    nrow(tcga_model_data),
    " TCGA patients and the external-validation cohort included ",
    nrow(cptac_model_data),
    " CPTAC patients. The final LASSO model retained ",
    number_selected,
    " non-zero gene-expression features. ",
    "The internally cross-validated TCGA AUC was ",
    tcga_auc_value,
    ", while the independent CPTAC AUC was ",
    cptac_auc_value,
    ". At the prespecified probability threshold of 0.5, ",
    cptac_predicted_methylated,
    " of ",
    length(cptac_class),
    " CPTAC patients were classified as methylated, giving a specificity of ",
    cptac_specificity_value,
    "."
  )
)
```

# 9. Interpretation

The model demonstrated modest discrimination in the independent CPTAC cohort
(AUC = 0.682), suggesting that some gene-expression signal associated with
the derived MGMT methylation outcome transferred between datasets. However,
at the prespecified probability threshold of 0.5, all CPTAC patients were
classified as methylated. This produced 100% sensitivity but 0% specificity.
The apparent accuracy of 62.9% therefore reflects the prevalence of the
methylated class rather than effective binary classification.

These findings provide evidence of limited cross-dataset ranking ability, but
not of a clinically useful or adequately calibrated binary predictor.
Differences in expression quantification, methylation measurement, outcome
construction, class balance and cohort composition may have contributed to
the poor calibration.

The model deliberately used genes measured in both datasets rather than every feature available in TCGA. This may reduce maximum within-cohort performance, but it increases portability and directly addresses the project brief's suggested compromise between data availability and data completeness.

# 10. Limitations

Several limitations should be considered:

1. The model predicts a temozolomide-response-associated biomarker, not
   observed patient response. MGMT methylation must therefore be interpreted
   as a clinically relevant molecular proxy rather than a direct
   treatment-response endpoint.
2. The TCGA and CPTAC expression datasets use different quantification pipelines.
3. The supplied CPTAC matrix does not contain CpG identifiers, so the
   methylation outcome is a platform-specific average across rows annotated
   to `MGMT`, classified using a TCGA-derived threshold, rather than
   MGMT-STP27 or a clinical laboratory assay.
4. CPTAC contains fewer patients than TCGA, which increases uncertainty around external performance estimates.
5. The classifier should not be interpreted as clinically deployable without further independent validation and calibration.

# 11. Conclusion

A gene-expression predictor developed in TCGA retained modest discriminatory
information when applied to CPTAC, but it did not generalise successfully as
a binary classifier at the prespecified threshold. The analysis therefore
supports modest cross-dataset discrimination but demonstrates inadequate
cross-dataset calibration.

Further harmonisation, calibration in a separate dataset and validation
against a clinically established MGMT outcome would be required before the
model could be considered practically or clinically useful. A threshold
selected using CPTAC outcomes would require evaluation in a third independent
cohort and is not presented here as validated performance.
